---
title: "Outbreak at a Convention: Epidemiologic Investigation"
author: "Epidemiology Class Discussion"
date: "2025-04-01"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo
editor: visual
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(readr)
library(epitools)
```

## Question 1: Suspected Agents and Incubation Periods

Staphylococcus aureus is suspected due to the rapid onset (2–6 hours) of symptoms: nausea, vomiting, diarrhea, and cramps.

## Question 2: Why Consider Incubation Period?

To narrow down potential pathogens and identify likely exposure windows.

## Question 3: Next Steps in Outbreak Investigation

1.  Confirm outbreak
2.  Define and identify cases
3.  Describe by person/place/time
4.  Develop hypotheses
5.  Evaluate hypotheses
6.  Refine hypotheses
7.  Implement control and prevention
8.  Communicate findings

## Question 4: Environmental Health Actions

-   Inspect kitchen and buffet
-   Assess food handling practices
-   Check temperatures of food storage
-   Test leftovers

## Question 5: Identifying the Food Source

Compare illness frequency among those who ate each buffet item vs. those who didn’t (cohort design).

## Question 6: Study Design

A **cohort study** is best since we have a defined population (attendees) and complete contact info.

## Question 7: Communication Method

Primary: Email questionnaire; Secondary: Phone follow-up.

## Question 8: Questionnaire Information to Collect

-   Demographics
-   Illness symptoms & onset
-   Foods consumed at buffet
-   Medical attention received

## Questionnaire Draft

```{markdown}
Dear Conference Attendee,

You are receiving this questionnaire because you attended the Museum Curators Conference on August 24th. Several attendees have become ill, and we are conducting a public health investigation.

Please take 5 minutes to complete the form below.

1. Name:
2. Email:
3. Sex:
4. Date of Birth:
5. Did you attend the lunch buffet on August 24? (Y/N)
6. What time did you eat lunch?
7. Which of the following did you eat? (Check all that apply)
   - Pasta primavera
   - Salad
   - Chicken
   - Bread rolls
   - Dessert
8. Did you experience any of the following symptoms?
   - Nausea
   - Vomiting
   - Diarrhea
   - Abdominal pain
9. When did symptoms begin (time/date)?
10. Did you seek medical care? (Y/N)

Thank you for your help!
```

## Line Listing of 5 Case Patients

```{r}
line_data <- read_csv("/Users/anapoleon/Documents/DS400/DS303_SP25/data/line_listing.csv")
line_data

epi_data <- read_csv("/Users/anapoleon/Documents/DS400/DS303_SP25/data/epi_curve.csv")
epi_data

#/Users/anapoleon/Documents/DS400/DS303_SP25/data
```

## Epi Curve of Outbreak

```{r}

epi_data <- epi_data %>% 
  pivot_longer(cols = c('Suspect Cases', 'Confirmed Cases', 'Total Cases'),
               names_to = "case_type",
               values_to = "number_of_cases")

ggplot(epi_data, aes(x = Hour, y = number_of_cases, color = case_type)) +
  geom_line(size = 1.2) +
  labs(title = "Epidemic Case Trends Over Time",
       x = "Date",
       y = "Number of Cases",
       color = "Case Type") +
  theme_minimal()
```

## Question 10: Point Source or Propagated?

**Point source outbreak** — all cases occurred over a short period after a common exposure (lunch buffet).

## Question 11: Recommendations for Kitchen Staff

-   Ensure hot food is held above 140°F (60°C)
-   Improve staff hygiene (handwashing, gloves)
-   Repair or replace faulty food warmers
-   Limit time food sits out

## Question 12: Steps in an Outbreak Investigation

1.  Prepare for fieldwork
2.  Establish outbreak existence
3.  Verify diagnosis
4.  Define and identify cases
5.  Describe and orient data
6.  Develop hypotheses
7.  Evaluate hypotheses
8.  Refine hypotheses and do more studies
9.  Implement control/prevention
10. Communicate findings
