---
title: "Assignment 2: Wrapping up regression"
author: 
  - "Alii Napoleon"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/28/25
format: html
toc: true
editor: visual
theme: spacelab
---

# Assignment 2: Wrapping up linear and logistic regression\*

Assignment 2 covers linear and logistic regression models and includes many of the topics we have covered over the entire semester. You will be using `palmerpenguins` data, which includes a nice variety of continuous and categorical predictors.

To submit this assignment, render this file and save as a pdf. Upload the pdf to Canvas.

\*This assignment does not include mixed effects models, but you will be seeing that on another assignment.

## Load Packages

```{r, message = FALSE}
library(lme4)
library(tidyverse)
# install.packages("palmerpenguins") # install if needed
library(palmerpenguins)
```

## Load Data

Load the `penguins` data and examine it below (use `summary()`, etc.)

```{r}
penguins <- penguins


```

## Question 1: Describe the data

What data is contained in this data set? Describe at least four variables (excluding `year`), including what they represent and their data type. Lastly, describe whether you think `year` would be a useful predictor in this data.

-    The data contains information on three penguins species from the Palmer Archipelago: Aldélie, Gentoo, and Chinstrap. Variables describe physical attributes, as named, and include bill_length, bill_depth, flipper_length, and body-mass. All four variables present categorical data.

## Question 2: EDA

Explore your data visually. Create at least two visualizations that show the relationship between `flipper_length_mm` and its potential predictors.

```{r, warning = FALSE}

# Scatterplot: Flipper Length and Body Mass

ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(aes(color = species)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Flipper Length vs. Body Mass",
    x = "Body Mass (g)",
    y = "Flipper Length (mm)"
  )

# Scatterplot: Flipper Length and Bill Length

ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm)) +
  geom_point(aes(color = species)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Flipper Length vs. Bill Length",
    x = "Bill Length (mm)",
    y = "Flipper Length (mm)"
  )

```

## Question 3: Apply a linear regression

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g`. Interpret the slope and intercept.

```{r}
# Simple Linear Regression Model

model1 <- lm(flipper_length_mm ~ body_mass_g, data = penguins)

summary(model1)

```

Interpret your model output in your own words below. Be sure to include a sentence explaining how `body_mass_g` impacts `flipper_length_mm` and whether or not the effect is significant.

-   The results indicate that for every additional gram of body mass, flipper length increases by 0.0153 mm. At *p* \< 2e-16, the relationship is highly significant. The intercept, 136.7**,** represents the predicted flipper length, if it were possible to have a zero body mass.

## Question 4: Apply a multiple linear regression

Fit a linear regression model predicting `flipper_length_mm` from both `body_mass_g` and `bill_length_mm`. Interpret the slopes and intercept.

```{r}
# Multiple Linear Regression Model

model2 <- lm(flipper_length_mm ~ body_mass_g + bill_length_mm, data = penguins)

summary(model2)
```

Similar to Question 3, interpret the model output in your own words here:

-   The model indicates that flipper length increases by 0.0131 mm for each additional gram of body mass, assuming bill length stays the same. It also increases by 0.549 mm for each additional millimeter of bill length, assuming body mass stays the same. The 122.0 mm intercept represents the predicted flipper length, if both predictors were ever to be zero.

## Question 5: Include an interaction

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g,` `bill_length_mm`, and the interaction of the two. Interpret the slopes and intercept.

```{r}
# Linear Regression Model with Interaction

model3 <- lm(flipper_length_mm ~ body_mass_g * bill_length_mm, data = penguins)

summary(model3)


```

Interpret the model output in your own words below. If there was a change in the pattern of significance, try to explain the logic below as well.

-   The model shows that the relationship between body mass and flipper length depends on bill length, *p* = 0.00234. While neither body mass nor bill length are statistically significant on their own, their combined effect is. This suggests that the influence of one depends on the level of the other. The 170.6 mmintercept is the predicted flipper length, if both predictors were to ever to be zero.

## Question 6: Compare models

Compare the models you built in Questions 4 and 5 using `anova()`.

```{r}
# Anova Comparison of Models 2 and 3

anova(model2, model3)


```

-   Which is the better model? How do you know?

    -    Model 2 is better, because it shows greater improvement fit, *p* = 0.00234. The added interaction explains additional variation in flipper length that the simpler model misses.

-   Is it possible to compare the models from Questions 3 and 5 using the same method? Why or why not?

    -   No, because they are not nested and model 3 includes different predictors entirely. Anova only works where one model is a simpler version of the other.

## Question 7: Categorical predictors

Build a linear model that includes a categorical predictor of your choice. It is fine to stick with dummy coding. Optional: apply a different coding scheme AND interpret the output correctly for +1 extra credit.

```{r}
# Linear Model with 'Species' as a Predictor

model4 <- lm(flipper_length_mm ~ species, data = penguins)

summary(model4)

```

```{r}

# EXTRA CREDIT!!

# Applying a Sum Contrast

contrasts(penguins$species) <- contr.sum(3)  

colnames(contrasts(penguins$species)) <- c("Chinstrap_vs_avg", "Gentoo_vs_avg")

# Fitting the Model

model4_effects <- lm(flipper_length_mm ~ species, data = penguins)

summary(model4_effects)

```

-   What is the reference level of your categorical predictor?

    -   Using dummy coding in the first model, the reference level is Adélie, and all coefficients are interpreted in comparison to it. In the second model, using effects coding, there is no single reference level and the coefficients reflect differences from the overall species average.

-   What is your interpretation of this model output? Address all coefficients.

    -   In the dummy-coded model, Adélie penguins have an average flipper length of 189.95 mm. Chinstrap penguins have flippers that are 5.87 mm longer than Adelies, and Gentoo are 27.23 mm longer, both statistically significant differences.\

        In the effects-coded model, the average flipper length across all species is 200.99 mm. Chinstrap penguins are 11.03 mm below this average, and Gentoo are 5.16 mm below it—indicating that Adélie penguins are above average by the same amount needed to balance the differences.

## Question 8: Relevel your categorical variable

Relevel your categorical variable so that a **different** level becomes the reference. Then, run the same model you did in Question 7 and interpret the output.

Relevel:

```{r}
# Relevelling with Gentoo as the Reference Category

penguins$species <- relevel(penguins$species, ref = "Gentoo")

```

Apply model from Question 7:

```{r}
# Refitting the Model

model5 <- lm(flipper_length_mm ~ species, data = penguins)

summary(model5)

```

-   What is the new reference level of your categorical predictor?

    -   Gentoo

-   What is your interpretation of this new model output? Address all coefficients.

    -   The differences below show that penguin species have different flipper lengths:
        -   Gentoo have flippers that are 217 mm long, on average
        -   Adélie are approximately 27 mm shorter
        -   Chinstrap are approximately 21 mm shorter than Gentoo

## Question 9: Apply a logistic regression

Apply a logistic regression. Include as many predictor variables as you'd like. Remember that your predicted outcome variable needs to be binary (or categorical with two levels).

\
Hint: You could use `sex` or create a binary variable of your own (e.g., Gentoo vs. non-Gentoo) to test your model.

```{r}
# Create a Binary Variable: TRUE = Gentoo, FALSE = otherwise

penguins$gentoo_flag <- ifelse(penguins$species == "Gentoo", 1, 0)

# Model with 'sex' as Added Predictor

logit_model2 <- glm(gentoo_flag ~ body_mass_g + flipper_length_mm + sex,
                    data = penguins,
                    family = "binomial")

summary(logit_model2)

```

-   What are your key takeaways from this model?

    -   The model broke, because it predicted the outcome too perfectly, causing errors. The results are unreliable, and none of the predictors show a clear or trustworthy effect.

## Question 10: Synthesize the information

-   Imagine you’re a biologist studying penguin populations. Which predictors do you think are most important to measure or record in the field to predict flipper length? Why?

    -   I would prioritize body mass and bill length, because they showed consistent relation to flipper lengths.

## Bonus: Stepwise Regression

Perform stepwise regression to find the best model for an outcome of your choice. You will likely encounter an error – fixing that error and explaining your findings will earn you +1 extra credit. Show your work.

```{r}


```

According to this stepwise regression, explain how the final model was selected.

**Answer:** (Write answer here)
